---
title: 'PARADIGM example: Figure 5'
author: "Chi L. Nguyen"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = c('png', 'postscript'))
library(tidyverse)
library(ggrepel)
library(ggpubr)
'%notin%' = Negate('%in%')
```

Table details:  
'tbldrugs' List of drug exposures and date of exposures for the entire MSKCC patient cohort (beween day -14 to 14 relative to HCT).  
'tblantibiotics_duke' List of antibiotics and date of exposures for the Duke patient cohort (between day -14 to 14 relative to HCT).  
'tblsample' Sample characteristics.  
'tblpatient' Patient charactestics, including discovery and validation set stratification.  
'tblresponse_score' Response scores for 4 microbiome features of interest. 

```{r load_data}

tbldrugs = read.csv("../../data/tbldrugs_MSKCC.csv")
head(tbldrugs)
length(unique(tbldrugs$PatientID))
range(tbldrugs$exposure_day_relative_to_hct)

tblantibiotics_duke = read.csv("../../data/tblantibiotics_Duke.csv")
head(tblantibiotics_duke)

tblsample = read.csv("../../data/tblsample.csv")
dim(tblsample)

tblpatient = read.csv("../../data/tblpatient.csv")
table(tblpatient$set)
head(tblpatient)

tblpatient_validation = tblpatient %>% 
  filter(set == "validation") 
length(unique(tblpatient_validation$PatientID))

tblresponse_score = read.csv("../../data/tblresponse_scores_4features_PARADIGM.csv")
head(tblresponse_score)

```

Pre-specified parameters. In this study, we investigated drug exposures between day -14 to 14 relative to HCT, and their associations with microbiome trajectories following drug exposures up until a month afterwards (day 14 to 45 relative to HCT). 

```{r parameter_setting}

drug_timepoint = -14:14 
microbiome_timepoint = 14:45 

```

We first explored *Enterococcus*-specific patient response scores. A negative score indicates that a given patient's drug exposure profiles between day -14 to 14 relative to HCT is associated with *Enterococcus* inhibition, and a positive score indicates that a given patient's drug exposure profiles between day -14 to 14 relative to HCT is associated with *Enterococcus* expansion. 
 
```{r Fig5b}

patient_level_response_score_Enterococcus = tbldrugs %>% 
    left_join(tblresponse_score %>% select(exposure_name, all_of("Enterococcus") ), by = "exposure_name") %>% 
    filter(exposure_day_relative_to_hct %in% drug_timepoint) %>%
    group_by(PatientID) %>%
    summarise(feature_name_rc = across(all_of("Enterococcus"), as.numeric)) %>% 
    summarise(rc = sum(feature_name_rc)) %>% 
    mutate(rc_norm = as.numeric( scale(rc) ) ) %>% 
    filter(PatientID %in% tblpatient_validation$PatientID) 

gghistogram(data = patient_level_response_score_Enterococcus, x = "rc_norm") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_y_continuous(limits = c(0,60), breaks = seq(0,60,by=10) ) +
  xlab("Patient-specific Enterococcus response scores") + 
  ylab("Number of patients") +
  geom_segment(aes(x = -0.5, y = 55, xend = -2, yend = 55), arrow = arrow(length = unit(0.2, "cm")), 
               color = "light blue") + 
  annotate(geom = "text", x = -1.3, y = 55.2, label = "Enterococcus-inhibiting\n association", 
           color = "light blue", size = 4) +
  geom_segment(aes(x = 1, y = 55, xend = 2.55, yend = 55), arrow = arrow(length = unit(0.2, "cm")), 
               color = "dark blue") +
  annotate(geom = "text", x = 1.7, y = 55.2, label = "Enterococcus-promoting\n association", 
           color = "dark blue", size = 4) 

```

Let's explore whether *Enterococcus*-specific patient response scores (based solely on drug exposures between day -14 to 14 relative to HCT) are predictive of future microbiome trajectories (based on stool samples collected between day 14 to 45 relative to HCT). 

```{r genus_abundance_prediction_Enterococcus_example}

genus_abundance_correlation_Enterococcus = tblsample %>% 
    filter(day_relative_to_hct %in% microbiome_timepoint) %>% 
    group_by(PatientID) %>% 
    summarise(genus_abundance = as.numeric( across(all_of("Enterococcus"), median)) ) %>% 
    select(PatientID, genus_abundance) %>% 
    filter(PatientID %in% patient_level_response_score_Enterococcus$PatientID) %>% 
    left_join(patient_level_response_score_Enterococcus, by = "PatientID") %>% 
    mutate(feature_name = "Enterococcus")

ggplot(data = genus_abundance_correlation_Enterococcus, aes(x=rc_norm, y = log10(genus_abundance+0.0001), fill = feature_name, col = feature_name)) +
  geom_point() + theme_classic() + geom_smooth(method = "lm") + stat_cor() +
  scale_color_manual(values = c("#129246"), name = "Feature name") +
  scale_fill_manual(values = c("#129246"), name = "Feature name") + 
  scale_y_continuous(limits = c(-4,1)) +
  xlab("Patient-specific\n bacteria response scores") +
  ylab("Genus abundance OR\n Alpha-diversity")  

```

To reproduce Fig 5c, we looped the script through all four features of interest. 

```{r patient_response_score_4features}

list_of_microbiome_features = c("Blautia", "Enterococcus", "Erysipelatoclostridium", "simpson_reciprocal")
genus_abundance_aggregate = data.frame()
patient_level_response_score_aggregate = data.frame(PatientID = tblpatient_validation$PatientID)

for (m in list_of_microbiome_features){
  
  patient_level_response_score = tbldrugs %>% 
    filter(PatientID %in% tblpatient_validation$PatientID) %>% 
    left_join(tblresponse_score %>% select(exposure_name, all_of(m) ), by = "exposure_name") %>% 
    filter(exposure_day_relative_to_hct %in% drug_timepoint) %>%
    group_by(PatientID) %>%
    summarise(feature_name_rc = across(all_of(m), as.numeric)) %>% 
    summarise(rc = sum(feature_name_rc)) %>% 
    mutate(rc_norm = as.numeric( scale(rc) ) ) 
  
  patient_level_response_score_aggregate = patient_level_response_score_aggregate %>% 
    left_join(patient_level_response_score, by = "PatientID")
  
  genus_abundance_correlation = tblsample %>%
    filter(PatientID %in% patient_level_response_score$PatientID) %>% 
    filter(day_relative_to_hct %in% microbiome_timepoint) %>% 
    group_by(PatientID) %>% 
    summarise(genus_abundance = as.numeric( across(all_of(m), median)) ) %>% 
    select(PatientID, genus_abundance) %>% 
    left_join(patient_level_response_score, by = "PatientID") %>% 
    mutate(feature_name = m)
  
  print(m)
  print(length(unique(patient_level_response_score$PatientID)))
  print(length(unique(genus_abundance_correlation$PatientID)))
  
  genus_abundance_aggregate = rbind.data.frame(genus_abundance_aggregate, genus_abundance_correlation)
  
}

colnames(patient_level_response_score_aggregate) = c("PatientID", "Blautia_rc", "Blautia_rc_norm",
                                                     "Enterococcus_rc", "Enterococcus_rc_norm", 
                                                     "Erysipelatoclostridium_rc", "Erysipelatoclostridium_rc_norm",
                                                     "simpson_reciprocal_rc", "simpson_reciprocal_rc_norm")

```

```{r Fig5c_MSKCC}

color_genus_abundance = c("#EC9B96", "#129246", "#7A6920", "black")
names(color_genus_abundance) = list_of_microbiome_features

ggplot(data = genus_abundance_aggregate, aes(x=rc_norm, y = log10(genus_abundance+0.0001), fill = feature_name, col = feature_name)) +
  geom_point() + theme_classic() + geom_smooth(method = "lm") + stat_cor() +
  scale_color_manual(values = color_genus_abundance, name = "Feature name") +
  scale_fill_manual(values = color_genus_abundance, name = "Feature name") + 
  scale_y_continuous(limits = c(-5,2)) +
  xlab("Patient-specific\n bacteria response scores") +
  ylab("Genus abundance OR\n Alpha-diversity")  

```

Similarly, we could calculate the patient response scores for the Duke validation cohort. For this cohort, we only evaluated antibiotics exposures. 

```{r Fig5c_Duke}

tblpatient_duke = tblpatient %>% 
  filter(set == "validation_Duke") 
length(unique(tblpatient_duke$PatientID))

genus_abundance_aggregate_duke = data.frame()
patient_level_response_score_aggregate_duke = data.frame(PatientID = tblpatient_duke$PatientID)

for (m in list_of_microbiome_features){
  
  patient_level_response_score = tblantibiotics_duke %>% 
    filter(PatientID %in% tblpatient_duke$PatientID) %>%
    left_join(tblresponse_score %>% select(exposure_name, all_of(m) ), by = "exposure_name") %>% 
    filter(exposure_day_relative_to_hct %in% drug_timepoint) %>%
    group_by(PatientID) %>%
    summarise(feature_name_rc = across(all_of(m), as.numeric)) %>% 
    summarise(rc = sum(feature_name_rc)) %>% 
    mutate(rc_norm = as.numeric( scale(rc) ) )
  
  patient_level_response_score_aggregate_duke = patient_level_response_score_aggregate_duke %>% 
    left_join(patient_level_response_score, by = "PatientID")
  
  genus_abundance_correlation = tblsample %>% 
    filter(PatientID %in% patient_level_response_score$PatientID) %>% 
    filter(day_relative_to_hct %in% microbiome_timepoint) %>% 
    group_by(PatientID) %>% 
    summarise(genus_abundance = as.numeric( across(all_of(m), median)) ) %>% 
    select(PatientID, genus_abundance) %>% 
    left_join(patient_level_response_score, by = "PatientID") %>% 
    mutate(feature_name = m)
  
  print(m)
  print(length(unique(patient_level_response_score$PatientID)))
  print(length(unique(genus_abundance_correlation$PatientID)))
  
  genus_abundance_aggregate_duke = rbind.data.frame(genus_abundance_aggregate_duke, genus_abundance_correlation)
  
}

colnames(patient_level_response_score_aggregate_duke) = c("PatientID", "Blautia_rc", "Blautia_rc_norm",
                                                     "Enterococcus_rc", "Enterococcus_rc_norm", 
                                                     "Erysipelatoclostridium_rc", "Erysipelatoclostridium_rc_norm",
                                                     "simpson_reciprocal_rc", "simpson_reciprocal_rc_norm")

ggplot(data = genus_abundance_aggregate_duke, aes(x=rc_norm, y = log10(genus_abundance+0.0001), fill = feature_name, col = feature_name)) +
  geom_point() + theme_classic() + geom_smooth(method = "lm") + stat_cor() +
  scale_color_manual(values = color_genus_abundance, name = "Feature name") +
  scale_fill_manual(values = color_genus_abundance, name = "Feature name") + 
  scale_y_continuous(limits = c(-5,2)) +
  xlab("Patient-specific\n bacteria response scores") +
  ylab("Genus abundance OR\n Alpha-diversity")  

```
