---
title: 'PARADIGM example: Figure 2h- cluster transition network'
author: "Chi L. Nguyen"
date: "10/17/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(igraph)
library(pheatmap)
```

Cohort characteristics are outlined in 'tblsample' table.

```{r, echo=TRUE}

tblsample = read.csv("~/Desktop/Backup from server /backup_Oct2020/First rotation - VDB/Data - ASV/Sept7_rebuttal/deposited dataset/tblsample_cohort_master_table_deid_MSKCC_9674_Duke_500_101222.csv")

```

We created sample-pair-wise table for patients with longitudinal sampling in the MSKCC discovery cohort. 

```{r, echo=TRUE}

cluster_on_t_end <- function(taxonomic_table){
  return(cbind.data.frame(oligos_id = taxonomic_table$oligos_id, subsequent_cluster = c(taxonomic_table$cluster_assignment[-1],NA), 
                          day.y = c(taxonomic_table$day_relative_to_hct[-1],NA)) )
}

subsequent_cluster = tblsample %>%
  filter(set == "discovery") %>% 
  arrange(PatientID) %>% 
  arrange(day_relative_to_hct) %>% 
  group_split(PatientID) %>%
  map_dfr(~cluster_on_t_end(.))
head(subsequent_cluster)

pairing_samples = tblsample %>% 
  left_join(subsequent_cluster, by = "oligos_id") %>% 
  filter(!is.na(subsequent_cluster)) %>% 
  mutate(dday = day.y - day_relative_to_hct) %>% 
  filter(dday <= 7) %>% 
  filter(set == "discovery")
head(pairing_samples)

```

We then calculated the cluster transition frequency among subsequently collected samples. 

```{r, echo=TRUE}

transition_matrix = table(pairing_samples$cluster_assignment, pairing_samples$subsequent_cluster)
transition_matrix = transition_matrix/rowSums(transition_matrix)
transition_matrix
```

We could visualize cluster transitions in a network map. 

```{r, echo=TRUE}

colors_asv_beta <- c("#A50026","#D73027","#F46D43","#C2A5CF","#3690C0",
                     "#FDAE61","#5AAE61","#B3E2CD","#762A83","#1B7837")

g <- graph.adjacency(transition_matrix, mode="directed", weighted=TRUE)
val2rgb <- colorRamp(rev(RColorBrewer::brewer.pal(11, "RdYlBu")[c(1:4,8:10)]), bias = 1)
size = table(tblsample$cluster_assignment[tblsample$set == "discovery"])
edge.curve = c(0.15)
layout <- matrix(c(1,5, -0.5,2, 1.5,1.5, 0.55,0.3, 2,0.65, 3,1, 3,2, 3,3, 6,6, 7,7), nrow=10, ncol=2, byrow=T)

edge.colored = rep(NA, 100)
edge.colored[which(E(g)$weight < 0.1)] <- 'grey'
edge.colored[which(E(g)$weight >= 0.1)] <- rgb(val2rgb(E(g)$weight[which(E(g)$weight >= 0.1)]) / 255)

edge.widthed = rep(NA,100)
edge.widthed[which(edge.colored == "grey")] <- 0.2
edge.widthed[which(is.na(edge.widthed))] <- E(g)$weight[which(is.na(edge.widthed))] * 8

plot.igraph(
  x               = g,
  xlim            = c(-1.5, 1.5),
  ylim            = c(-1.5, 1.5),
  layout          = layout_in_circle,
  vertex.size     = sqrt(size), vertex.color = colors_asv_beta, edge.arrow.size = 0.25,
  edge.width      = edge.widthed,
  edge.color      = edge.colored,
  edge.curved     = edge.curve, 
  edge.loop.angle = scales::rescale(seq_along(E(g)), to=c(0, -2 * 3.14)) )

legend(
  x      = 1.5, 
  y      = 1, 
  fill   = c("grey", "#679DC9", "#FDAB60", "#DC3A2C"), 
  legend = c("<= 10%", "25%", "50%", "80%"), 
  title  = "Transition\nFrequency", 
  bty    = "n" )

```

We could also visualize cluster transitions in a 10x10 heatmap. 

```{r, echo=TRUE}

pheatmap((transition_matrix*100),cluster_rows= FALSE, cluster_cols = FALSE,
         display_numbers = round(transition_matrix,2), breaks = seq(0,100,by=1))

```