---
title: 'PARADIGM example: Figure 3b-c'
author: "Chi L. Nguyen"
date: "10/17/2022"
output: "github_document"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pheatmap)
source("~/Desktop/Backup from server /backup_Oct2020/First rotation - VDB/Script/github_deposit/paradigm_drug_cluster_association_v6.R", echo=F)

```

Source script runs elastic net regularized regression to learn the coefficients of each drug exposure to cluster self and attractor transitions, then convert cluster-drug associations into drug-taxon response scores. 

Table details: 
'tbldaily' Training data of microbiome cluster dynamics.
'tblpatient' Patient characteristics.
'tbleuclidean_distance' A 10x10 matrix of pair-wise cluster euclidean distance.
'tblcounts' Counts and taxonomic classifications of ASVs.
'tblsample' Sample characteristics. 

```{r load data}

tbldaily = read.csv("~/Desktop/Backup from server /backup_Oct2020/First rotation - VDB/Data - ASV/Sept7_rebuttal/deposited dataset/tbldaily_sampling_PARADIGM_input_101122.csv")
head(tbldaily)

tblpatient = read.csv("~/Desktop/Backup from server /backup_Oct2020/First rotation - VDB/Data - ASV/Sept7_rebuttal/deposited dataset/tblpatient_cohort_characteristics_master_table_deid_MSKCC_1227_Duke_142.csv")
head(tblpatient)

tbleuclidean_distance = read.csv("~/Desktop/Backup from server /backup_Oct2020/First rotation - VDB/Data - ASV/Sept7_rebuttal/deposited dataset/tbleuclidean_distance_10clusters.csv")
head(tbleuclidean_distance)

tblcounts = read.csv("~/Desktop/Backup from server /backup_Oct2020/First rotation - VDB/Data - ASV/Sept7_rebuttal/deposited dataset/tblcounts_master_table_deid_MSKCC_9674_Duke_500_101222.csv")
head(tblcounts)

tblsample = read.csv("~/Desktop/Backup from server /backup_Oct2020/First rotation - VDB/Data - ASV/Sept7_rebuttal/deposited dataset/tblsample_cohort_master_table_deid_MSKCC_9674_Duke_500_101222.csv")
head(tblsample)

```

We created two data frames, one containing the response variable (cluster self/attractor transtions) and one containing the predictors (time of sample collection and drug exposures). 

```{r create_response_matrices}

cluster_dynamic = tbldaily[,1:5]
head(cluster_dynamic) 

environmental_exposures = tbldaily[,c(6,9:ncol(tbldaily))]
head(environmental_exposures)

```

We calculated coefficients of predictors to cluster self and attractor transitions.
Expected output is a matrix, with predictors column-wise and coefficient values row-wise. 
The parameters are explained in the source script. 

```{r calculate_self_attractor_coefficients}

self_coefficient_matrix = self_transition(cluster_dynamic = cluster_dynamic, 
                                          environmental_exposures = environmental_exposures, 
                                          seed = 2208, 
                                          numFold = 10, 
                                          numCluster = 10)
head(self_coefficient_matrix)

attractor_coefficient_matrix = attractor_transition(cluster_dynamic = cluster_dynamic, 
                                                    environmental_exposures = environmental_exposures,
                                                    euclidean_distance = tbleuclidean_distance, 
                                                    seed = 2208, 
                                                    numFold = 10, 
                                                    numCluster = 10)

head(attractor_coefficient_matrix)

```

To plot Fig. 3b and Fig.S3, we arranged drug exposures by drug class. 

```{r drug_classification}

self_coefficient_to_plot = t(self_coefficient_matrix)
attractor_coefficient_to_plot = t(attractor_coefficient_matrix)

tblgraph_drug_exposure_classification = read.csv("~/Desktop/Backup from server /backup_Oct2020/First rotation - VDB/Data - ASV/Sept7_rebuttal/deposited dataset/tblgraph_drug_exposure_classification.csv") 

drug_group_annotation = tblgraph_drug_exposure_classification %>% select(group, ind)
rownames(drug_group_annotation) = tblgraph_drug_exposure_classification$exposure_name

```

We could now plot a heatmap for the coefficients of the elastic net regularized regression model, which indicates the magnitude and direction of association between a given exposures and cluster self transitions...

```{r FigS3_self}

set = c("#FFFFFF", RColorBrewer::brewer.pal(9, "Set1"), RColorBrewer::brewer.pal(12, "Set3")[3:7])
set = list(group = set)
names(set$group) = unique(tblgraph_drug_exposure_classification$group)

ind_df = match(rownames(drug_group_annotation), rownames(self_coefficient_to_plot)) 

out_self = pheatmap(self_coefficient_to_plot[c(1,2,ind_df),], col = colorRampPalette(c("navy", "white", "firebrick3"), bias = 1.3)(150), 
                    annotation_row = drug_group_annotation, annotation_colors = set, 
                    cluster_cols = F, cluster_rows = F, breaks = seq(-3,4,by=0.05), main = "Self Transition")

```

And cluster attractor transitions. 

```{r FigS3_attractor}

out_attractor = pheatmap(attractor_coefficient_to_plot[c(1,2,ind_df),], col = colorRampPalette(c("navy", "white", "firebrick3"), bias = 1.3)(150), 
                         annotation_row = drug_group_annotation, annotation_colors = set,
                         cluster_cols = F, cluster_rows = F, breaks = seq(-3,4,by=0.05), main = "Attractor Transition")    

```

We trained the model and learned the associations between a drug exposure and a given cluster self and attractor transitions. A negative coefficient value indicates that a drug exposure is associated with decreased cluster self/attractor transition probability, and a positive coefficient value indicates that a drug exposure is associated with increased cluster self/attracotr transition probability. 

To convert drug-cluster associations into drug-taxon associations, we calculated bacteria response scores, which measure the magnitude and direction of drug-taxon associations. Deriving the bacteria response score depends on cluster transition probability matrix and cluster-specific mean measurement values of the taxonomic feature of interest. The source script contains a function to calculate cluster transition probability matrix based on the self and attractor coefficient matrix. 

First, we determined the mean taxonomic relative abundance (or alpha-diversity) per cluster. Users could explore other microbiome features of interest (if available in either the 'tblsample' or the 'tblcounts' tables). 

In our study, we focused on microbiome features that have been associated with allo-HCT outcomes, including relative abundances of genus Blautia, Enterococcus, Erysipelatoclostridium, and alpha-diversity (Simpson reciprocal). 

```{r calculate_cluster_feature_mean}

drugList = colnames(environmental_exposures)[-1]
microbiomeFeatureList = c("Blautia", "Enterococcus", "Erysipelatoclostridium", "simpson_reciprocal")
feature_profile_by_cluster = list()
for (m in microbiomeFeatureList){
  
  feature_profile = tblsample %>% 
    filter(set == "discovery") %>% 
    group_by(cluster_assignment) %>% 
    summarise(across(all_of(m), mean)) 
  
  feature_profile_by_cluster = append(feature_profile_by_cluster, list(feature_profile[[m]]) )
} 

```

We then calculated the bacteria response scores for 62 investigated medications across these four microbiome features. 

```{r calculate_response_scores}

response_score_feature_aggregate = data.frame(exposure_name = drugList)

for (i in 1:length(microbiomeFeatureList)){
  
  feature_profile_by_cluster_i = feature_profile_by_cluster[[i]]
  feature_profile_name = microbiomeFeatureList[i]
  
  response_score_aggregate = c() 
  
  for (drug in drugList){
    
    response_score = bacteria_response_score(self_coef = self_coefficient_matrix, 
                                             attractor_coef = attractor_coefficient_matrix, 
                                             exposure_name = drug,
                                             feature_profile_by_cluster = feature_profile_by_cluster_i, 
                                             euclidean_distance = tbleuclidean_distance, 
                                             d = 0, 
                                             numCluster = 10)
                                                  
    
    response_score_aggregate = c(response_score_aggregate, response_score)
                          
  }
  response_score_feature_aggregate[feature_profile_name] = response_score_aggregate
}

head(response_score_feature_aggregate)

```

The 'response_score_feature_aggregate' table contains bacteria response score values, with each row being a drug exposure and each column being a microbiome feature. A negative response score indicates that a drug exposure is associated with decreased taxon relative abundance (or alpha-diversity), and a positive response score indicates that a drug exposure is associated with increased taxon relative abundance (or alpha-diversity). 

```{r FigS4}

drug_level = rownames(drug_group_annotation)
response_score_feature_aggregate_to_plot = melt(response_score_feature_aggregate, id = "exposure_name")
response_score_feature_aggregate_to_plot = response_score_feature_aggregate_to_plot %>% 
  mutate(sign_coef = case_when(value < 0 ~ "-", 
                               TRUE ~ "+"))

ggplot(data = response_score_feature_aggregate_to_plot, 
       aes(x=variable, y=factor(exposure_name, levels = rev(drug_level)), fill=value )) +
  geom_tile(stat = "identity") + theme_classic() + 
  theme(plot.title=element_text(size=16,face="bold"),
        axis.title.y=element_blank(),
        axis.ticks.y=element_blank()) +
  geom_text(aes(label = sign_coef), color = "black", size = 4) +
  coord_flip() + coord_fixed() +
  scale_fill_gradientn(colours = colorRampPalette(c("navy", "white", "firebrick3"), bias = 1)(30)) + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.y = element_blank(), axis.title.x = element_blank())

```

As a sanity check, we correlated the bacteria response scores for alpha-diversity and for Enterococcus relative abundance. We expected and observed a significant negative correlation, indicating that drug exposures which are associated with increased Enterococcus relative abundnace, are associated with decreased alpha-diversity, and vice-versa. In allo-HCT, Enterococcus expansion typically leads to a low-diversity state of Enterococcus domination and dysbiosis. 

```{r Fig3d}

ggplot(data = response_score_feature_aggregate, 
       aes(x=Enterococcus, y=simpson_reciprocal)) +
  geom_point() +
  geom_smooth(method = "lm", linetype = "dashed") +
  stat_cor() +
  theme_classic() +
  xlab("Enterococcus \n response scores") +
  ylab("Alpha-diversity \n response scores")

```

